From Reversible Programs to \\ Univalent Universes and Back

### 2021-11-12 Pi/paper/applications.tex
Fix headlines and captions

\subsection{Synthesis and verification}

@@

\noindent
The permutation uses the canonical encoding of sequences of bits as natural numbers, e.g.,
${(\mathsf{false},\mathsf{true},\mathsf{true})}$ is encoded as 011 or 3. The second entry in the $\term{lookup}$ table
maps index 1 (= 001) to the value 5 (= 101), following the reversible disjunction specification (recall that
$\mathsf{reversibleOr}(h,b_1,b_2) = (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
and~$\underline{\vee}$ is exclusive-disjunction). Quoting this permutation generates the same normalised program
\Afun{reversibleOrNorm}, matching the desired structured type of a 1-combinator on $\mathbb{8}+$.

@@

\subsection{Reasoning}

@@

We define a function $\parity : (X \iso Y) \to \Bool$ to compute the parity of a permutation defined by a combinator
$c$, by first transforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$ defined
in~\cref{sec:equivalence}), and then computing the parity of this representation, which amounts to a simple checking of
the parity of the length of the list.

@@

\noindent
Using~\cref{prop:parity-preserved} and the fact that any composition of even circuits is even, we complete the proof by
computing the $\parity$ of \Afun{toffoli₄}, and all of the \Afun{toffoli₃} combinations.

>>>>

\subsection{Synthesis and Verification}

@@

\noindent
The permutation uses the canonical encoding of sequences of bits as natural numbers, e.g.,
${(\mathsf{false},\mathsf{true},\mathsf{true})}$ is encoded as 011 or 3. The second entry in the $\term{lookup}$ table
maps index 1 (${= 001}$) to the value 5 (${= 101}$), following the reversible disjunction specification (recall the definition
$\mathsf{reversibleOr}(h,b_1,b_2) = (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
and~$\underline{\vee}$ is exclusive-disjunction). Quoting this permutation generates the same normalised program
\Afun{reversibleOrNorm}, matching the desired structured type of a 1-combinator on $\mathbb{8}+$.

@@

\subsection{Reasoning and Transfer of Theorems}

@@

We define a function $\parity : (X \iso Y) \to \Bool$ to compute the parity of a permutation defined by a combinator
$c$, by first transforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$ from~\cref{def:evalh}),
and then computing the parity of this representation, which amounts to a simple checking of the parity of the length of
the list.

@@

\noindent
Using~\cref{prop:parity-preserved}, we complete the proof by computing the $\parity$ of \Afun{toffoli₄}, and all
combinations of the \Afun{toffoli₃} gates.

### 2021-11-12 Pi/paper/discussion.tex

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible computing based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was
suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{caretteEmbracingLawsPhysics2021}. In this work, we
have established this correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we
discuss some broader related work.

>>>>

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible computing based on algebraic types matches \emph{exactly} the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
Historically, the first such correspondence related the $\lambda$-calculus, intuitionistic logic, and cartesian-closed
categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard correspondence was
suggested by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was suggested
by~\citet*{caretteComputingSemiringsWeak2016} and~\citet*{caretteEmbracingLawsPhysics2021}. For a 1-bit fragment of
$\PiLang$, this was established by~\citet*{caretteReversibleProgramsUnivalent2018}. In this work, we have completely
established this correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we discuss
some broader related work.
